---
title: "MultiSplines"
output: html_notebook
---

DISAGIO USABLEVS30--> NON VA
RANDOM EFFECT è SIGNIFICATIVO --> COSA COMPORTA?

LIBRARIES and WD

Per rimuovere una covariata dal modello serve un permutational test se non ho gaussianità.

posso fare le confidence bands senza gaussianità, test 03/12? --> bootstrap or permutation for confidence bands

```{r}
#setwd("C:/Users/paolo/OneDrive/Desktop/Nonparam/progetto")

library(MASS)
library(rgl)
library(DepthProc)
library(gam)
library(plot3D)

```


I load my data and I remove last covariates already excluded from analysis (CHECK).
There should be a missing factor for clusterization along depth.
I apply a log transform!
MStat==ML!
Try factor station

```{r}
DATA<-read.csv("FAS_000_50_cleaned.csv", header=TRUE, sep=",")
colnames(DATA)
Station=factor(DATA$Stat)
attach(DATA)

```


Utility Function

```{r}

gamR2 <- function(gam){
R2 <- 1-((sum(residuals(gam)^2))/
(sum((gam$y - mean(gam$y))^2)))
R2adj <- 1- ((1 - R2) * (length(gam$y) - 1)/
(length(gam$y) - length(gam$coefficients) - 1))
a <- data.frame(R2, R2adj)
return(a)
}

ggplot.model <- function(model, type="conditional", res=FALSE,
                       col.line="#7fc97f", col.point="#beaed4", size.line=1, size.point=1) {
  require(visreg)
  require(plyr)
  plotdata <- visreg(model, type = type, plot = FALSE)
  smooths <- ldply(plotdata, function(part)
    data.frame(Variable = part$meta$x,
             x=part$fit[[part$meta$x]],
             smooth=part$fit$visregFit,
             lower=part$fit$visregLwr,
             upper=part$fit$visregUpr))
  residuals <- ldply(plotdata, function(part)
    data.frame(Variable = part$meta$x,
               x=part$res[[part$meta$x]],
               y=part$res$visregRes))
  if (res)
    ggplot(smooths, aes(x, smooth)) + geom_line(col=col.line, size=size.line) +
      geom_line(aes(y=lower), linetype="dashed", col=col.line, size=size.line) +
      geom_line(aes(y=upper), linetype="dashed", col=col.line, size=size.line) +
      geom_point(data = residuals, aes(x, y), col=col.point, size=size.point) +
      facet_grid(. ~ Variable, scales = "free_x")
  else
    ggplot(smooths, aes(x, smooth)) + geom_line(col=col.line, size=size.line) +
      geom_line(aes(y=lower), linetype="dashed", col=col.line, size=size.line) +
      geom_line(aes(y=upper), linetype="dashed", col=col.line, size=size.line) +
      facet_grid(. ~ Variable, scales = "free_x")
}

```


Now let's plot the variable relationship with LogY.
I can't identify noteworthy cutoffs or knot points clearly.

```{r}
plot(DATA$ML,DATA$LogY)
plot(DATA$EvDpt,DATA$LogY)
plot(DATA$Dipo,DATA$LogY)
plot(DATA$UsableVS30,DATA$LogY)

```

Therefore try evenly spaced knots for splines. Start with:


EVDPT
bs doesn't handle well boundaries, while ns performs much better.
On the whole I'd choose the smoothed one.

```{r}
DF=7
knots.EvDpt=quantile(DATA$EvDpt,probs=c(0.1,0.5,0.9))
boundary.knots.EvDpt=quantile(DATA$EvDpt,probs=c(0.05,0.95))
normal_EvDpt= ns(EvDpt,knots=knots.EvDpt,Boundary.knots=boundary.knots.EvDpt)

model_EvDpt=lm(LogY ~ ns(EvDpt,knots=knots.EvDpt,Boundary.knots=boundary.knots.EvDpt)) #natural splines
#model_EvDpt=lm(LogY ~ bs(EvDpt, degree=3,df=DF)) #classical splines
#model_EvDpt=lm(LogY ~ EvDpt)
EvDpt.grid=seq(range(EvDpt)[1],range(EvDpt)[2],by=0.005)
preds=predict(model_EvDpt,list(EvDpt=EvDpt.grid) ,se=T)
se.bands=cbind(preds$fit +2* preds$se.fit ,preds$fit -2* preds$se.fit)
plot(EvDpt ,LogY ,xlim=range(EvDpt.grid) ,cex =.5, col =" darkgrey " )
lines(EvDpt.grid,preds$fit ,lwd =2, col =" blue")
matlines(EvDpt.grid ,se.bands ,lwd =1, col =" blue",lty =3)
knots=attr(bs(EvDpt, degree=3,df=DF),'knots')


summary(model_EvDpt)

#SMOOTHING SPLINES

smooth_EvDpt=smooth.spline(EvDpt,LogY,cv=T)
plot(EvDpt ,LogY,cex =.5, col =" darkgrey ")
lines(smooth_EvDpt,col="blue",lwd=2)
smooth_EvDpt$lambda
smooth_EvDpt$df

model_EvDpt=smooth_EvDpt

library(mgcv)
model_EvDpt=mgcv::gam(LogY ~ s(EvDpt, bs='cr'))
preds=predict(model_EvDpt,list(EvDpt=EvDpt.grid) ,se=T)
se.bands=cbind(preds$fit +2* preds$se.fit ,preds$fit -2* preds$se.fit)
plot(EvDpt ,LogY ,xlim=range(EvDpt.grid) ,cex =.5, col =" darkgrey " )
lines(EvDpt.grid,preds$fit ,lwd =2, col =" blue")
matlines(EvDpt.grid ,se.bands ,lwd =1, col =" blue",lty =3)
detach("package:mgcv", unload=TRUE)
                          


```
USABLEVS30
Smoothed splines are not far from classical linear models. Why not?
Ns varies a bit more.
BOH
```{r}
DF=7
knots.UsableVS30=quantile(DATA$UsableVS30,probs=c(0.1,0.5,0.9))
boundary.knots.UsableVS30=quantile(DATA$UsableVS30,probs=c(0.05,0.95))
normal_UsableVS30= ns(UsableVS30,knots=knots.UsableVS30,Boundary.knots=boundary.knots.UsableVS30)

model_UsableVS30=lm(LogY ~ ns(UsableVS30,knots=knots.UsableVS30,Boundary.knots=boundary.knots.UsableVS30)) #natural splines
#model_UsableVS30=lm(LogY ~ bs(UsableVS30, degree=3,df=DF)) #classical splines
#model_UsableVS30=lm(LogY ~ UsableVS30)
UsableVS30.grid=seq(range(UsableVS30)[1],range(UsableVS30)[2],by=0.005)
preds=predict(model_UsableVS30,list(UsableVS30=UsableVS30.grid) ,se=T)
se.bands=cbind(preds$fit +2* preds$se.fit ,preds$fit -2* preds$se.fit)
plot(UsableVS30 ,LogY ,xlim=range(UsableVS30.grid) ,cex =.5, col =" darkgrey " )
lines(UsableVS30.grid,preds$fit ,lwd =2, col =" blue")
matlines(UsableVS30.grid ,se.bands ,lwd =1, col =" blue",lty =3)
knots=attr(bs(UsableVS30, degree=3,df=DF),'knots')


summary(model_UsableVS30)

#SMOOTHING SPLINES

smooth_UsableVS30=smooth.spline(UsableVS30,LogY,cv=T)
plot(UsableVS30 ,LogY,cex =.5, col =" darkgrey ")
lines(smooth_UsableVS30,col="blue",lwd=2)
smooth_UsableVS30$lambda
smooth_UsableVS30$df

smooth_UsableVS30=smooth.spline(UsableVS30,LogY,lambda=1e-2)
plot(UsableVS30 ,LogY,cex =.5, col =" darkgrey ")
lines(smooth_UsableVS30,col="blue",lwd=2)
smooth_UsableVS30$lambda
smooth_UsableVS30$df

```

DIPO
Smoothed is wobbly!
Ns works fine.
```{r}
DF=7
knots.Dipo=quantile(DATA$Dipo,probs=c(0.1,0.5,0.9))
boundary.knots.Dipo=quantile(DATA$Dipo,probs=c(0.05,0.95))
normal_Dipo= ns(Dipo,knots=knots.Dipo,Boundary.knots=boundary.knots.Dipo)

model_Dipo=lm(LogY ~ ns(Dipo,knots=knots.Dipo,Boundary.knots=boundary.knots.Dipo)) #natural splines
#model_Dipo=lm(LogY ~ bs(Dipo, degree=3,df=DF)) #classical splines
#model_Dipo=lm(LogY ~ Dipo)
Dipo.grid=seq(range(Dipo)[1],range(Dipo)[2],by=0.005)
preds=predict(model_Dipo,list(Dipo=Dipo.grid) ,se=T)
se.bands=cbind(preds$fit +2* preds$se.fit ,preds$fit -2* preds$se.fit)
plot(Dipo ,LogY ,xlim=range(Dipo.grid) ,cex =.5, col =" darkgrey " )
lines(Dipo.grid,preds$fit ,lwd =2, col =" blue")
matlines(Dipo.grid ,se.bands ,lwd =1, col =" blue",lty =3)
knots=attr(bs(Dipo, degree=3,df=DF),'knots')


summary(model_Dipo)

#SMOOTHING SPLINES


smooth_Dipo=smooth.spline(Dipo,LogY,lambda=1e-2)
plot(Dipo ,LogY,cex =.5, col =" darkgrey ")
lines(smooth_Dipo,col="blue",lwd=2)
smooth_Dipo$lambda
smooth_Dipo$df


```


ML
I'd choose ns, since smooth splines weakly suggest a weird behavior for high Magnitudes.
```{r}



DF=7
knots.ML=quantile(DATA$ML,probs=c(0.1,0.5,0.9))
boundary.knots.ML=quantile(DATA$ML,probs=c(0.05,0.95))
normal_ML= ns(ML,knots=knots.ML,Boundary.knots=boundary.knots.ML)

model_ML=lm(LogY ~ ns(ML,knots=knots.ML,Boundary.knots=boundary.knots.ML)) #natural splines
#model_ML=lm(LogY ~ bs(ML, degree=3,df=DF)) #classical splines
#model_ML=lm(LogY ~ ML)
ML.grid=seq(range(ML)[1],range(ML)[2],by=0.005)
preds=predict(model_ML,list(ML=ML.grid) ,se=T)
se.bands=cbind(preds$fit +2* preds$se.fit ,preds$fit -2* preds$se.fit)
plot(ML ,LogY ,xlim=range(ML.grid) ,cex =.5, col =" darkgrey " )
lines(ML.grid,preds$fit ,lwd =2, col =" blue")
matlines(ML.grid ,se.bands ,lwd =1, col =" blue",lty =3)
knots=attr(bs(ML, degree=3,df=DF),'knots')


summary(model_ML)

#SMOOTHING SPLINES

smooth_ML=smooth.spline(ML,LogY,cv=T)
plot(ML ,LogY,cex =.5, col =" darkgrey ")
lines(smooth_ML,col="blue",lwd=2)
smooth_ML$lambda
smooth_ML$df




```


Now combine splines with GAM.
I have options depending on s() and other smoothing splines functions.

EVDPT vs DIPO

```{r}


#random uses factor fox mixed linear effects
gam1=gam::gam(LogY ~ s(EvDpt ,df=7)+s(Dipo ,df=8))
#gam1=gam::gam(LogY ~ ns(EvDpt,knots=knots.EvDpt,Boundary.knots=boundary.knots.EvDpt)+
                #  ns(Dipo,knots=knots.Dipo,Boundary.knots=boundary.knots.Dipo)) #plor problems

#gam1=lm(LogY ~ ns(EvDpt,knots=knots.EvDpt,Boundary.knots=boundary.knots.EvDpt)+
       #           ns(Dipo,knots=knots.Dipo,Boundary.knots=boundary.knots.Dipo)) #bad fit



EvDpt.grid=seq(range(EvDpt)[1],range(EvDpt)[2],by=1)
Dipo.grid=seq(range(Dipo)[1],range(Dipo)[2],by=1)
summary(gam1)
plot(gam1)
grid=expand.grid(EvDpt.grid,Dipo.grid)
names(grid)=c('EvDpt','Dipo')

pred=predict(gam1,newdata = grid) 

persp3D(z = pred, theta = 120)
persp3d(EvDpt.grid,Dipo.grid,pred,col='blue')
points3d(EvDpt, Dipo, LogY, col = "red", size = 2, add=T)

```

ML vs USABLEVS30


```{r}

#random uses factor fox mixed linear effects
gam1=gam::gam(LogY ~ s(ML,df=8)+s(UsableVS30 ,df=8))

ML.grid=seq(range(ML)[1],range(ML)[2],by=1)
UsableVS30.grid=seq(range(UsableVS30)[1],range(UsableVS30)[2],by=1)
summary(gam1)
plot(gam1)
grid=expand.grid(ML.grid,UsableVS30.grid)
names(grid)=c('ML','UsableVS30')

pred=predict(gam1,newdata = grid) 

persp3D(z = pred, theta = 120)
persp3d(ML.grid,UsableVS30.grid,pred,col='blue')
points3d(ML,UsableVS30, LogY, col = "red", size = 2, add=T)

```

ML vs Dipo

```{r}
#gam1=gam(LogY ~ s(ML ,df=12)+s(Dipo ,df=12))
#random uses factor fox mixed linear effects
gam1=gam::gam(LogY ~ s(ML,df=12)+s(Dipo ,df=12))

ML.grid=seq(range(ML)[1],range(ML)[2],by=1)
Dipo.grid=seq(range(Dipo)[1],range(Dipo)[2],by=1)
summary(gam1)
plot(gam1)
grid=expand.grid(ML.grid,Dipo.grid)
names(grid)=c('ML','Dipo')

pred=predict(gam1,newdata = grid) 

persp3D(z = pred, theta = 120)
persp3d(ML.grid,Dipo.grid,pred,col='blue')
points3d(ML, Dipo, LogY, col = "red", size = 2, add=T)

```

ML vs EVDPT

```{r}
#gam1=gam(LogY ~ s(ML ,df=12)+s(Dipo ,df=12))
#random uses factor fox mixed linear effects
gam1=gam::gam(LogY ~ s(ML,df=12)+s(EvDpt ,df=12))

ML.grid=seq(range(ML)[1],range(ML)[2],by=1)
EvDpt.grid=seq(range(EvDpt)[1],range(EvDpt)[2],by=1)
summary(gam1)
plot(gam1)
grid=expand.grid(ML.grid,EvDpt.grid)
names(grid)=c('ML','EvDpt')

pred=predict(gam1,newdata = grid) 

persp3D(z = pred, theta = 120)
persp3d(ML.grid,EvDpt.grid,pred,col='blue')
points3d(ML, EvDpt, LogY, col = "red", size = 2, add=T)

```

EVDPT vs UsableVS30

```{r}

gam1=gam::gam(LogY ~ s(UsableVS30,df=10)+s(EvDpt ,df=10))

UsableVS30.grid=seq(range(UsableVS30)[1],range(UsableVS30)[2],by=1)
EvDpt.grid=seq(range(EvDpt)[1],range(EvDpt)[2],by=1)
summary(gam1)
plot(gam1)
grid=expand.grid(UsableVS30.grid,EvDpt.grid)
names(grid)=c('UsableVS30','EvDpt')

pred=predict(gam1,newdata = grid) 

persp3D(z = pred, theta = 120)
persp3d(UsableVS30.grid,EvDpt.grid,pred,col='blue')
points3d(UsableVS30, EvDpt, LogY, col = "red", size = 2, add=T)

```

USAblEVS30 VS DIPO

```{r}


#gam1=gam::gam(LogY ~ s(UsableVS30,df=8)+s(Dipo ,df=10))

gam1=gam::gam(LogY ~ s(UsableVS30,df=6, spar=1)+s(Dipo ,df=8,spar=0.1))

UsableVS30.grid=seq(range(UsableVS30)[1],range(UsableVS30)[2],by=1)
Dipo.grid=seq(range(Dipo)[1],range(Dipo)[2],by=1)
summary(gam1)
plot(gam1)
grid=expand.grid(UsableVS30.grid,Dipo.grid)
names(grid)=c('UsableVS30','Dipo')

pred=predict(gam1,newdata = grid) 

persp3D(z = pred, theta = 120)
persp3d(UsableVS30.grid,Dipo.grid,pred,col='blue')
points3d(UsableVS30, Dipo, LogY, col = "red", size = 2, add=T)




```

##NOW ALL TOGETHER
  s(UsableVS30,df=6, spar=1)



gam1=gam::gam(LogY ~ s(Dipo ,df=10)+s(ML,df=12)+s(EvDpt ,df=8)+  s(UsableVS30,df=6, spar=1))

summary.Gam(gam1)
plot(gam1)
gamR2(gam1)





gam2<-gam::gam(LogY ~ ns(EvDpt,knots=knots.EvDpt,Boundary.knots=boundary.knots.EvDpt)+
                ns(Dipo,knots=knots.Dipo,Boundary.knots=boundary.knots.Dipo)+
                 ns(ML,knots=knots.ML,Boundary.knots=boundary.knots.ML)
               )
summary(gam2)
gamR2(gam2)

print(anova(gam1,best_gam,test='F'))
print(anova(gam1,gam2,test='F'))
print(anova(best_gam,gam4,test='F'))


OTHER MODELS - Rimuovo usableVS30 ##non mi convince



gam3<-gam::gam(LogY ~ ns(EvDpt,knots=knots.EvDpt,Boundary.knots=boundary.knots.EvDpt)+
                ns(Dipo,knots=knots.Dipo,Boundary.knots=boundary.knots.Dipo)+
                 ns(ML,knots=knots.ML,Boundary.knots=boundary.knots.ML)
               )
summary(gam3)
gamR2(gam3)


gam4<-gam::gam(LogY ~ ns(EvDpt,knots=knots.EvDpt,Boundary.knots=boundary.knots.EvDpt)+
                ns(Dipo,knots=knots.Dipo,Boundary.knots=boundary.knots.Dipo)+
                 ns(ML,knots=knots.ML,Boundary.knots=boundary.knots.ML)
               +random(Station))
summary(gam4)
gamR2(gam4)


print(anova(gam2,gam3,test='F'))  #model 4 beats model 3, yet the p-value for UsableVS30 is much higher
print(anova(gam3,gam4,test='F'))  #model 5 > model 4
print(anova(best_gam,gam4,test='F'))  #model 5 > model 4

gam4$coefficients



PERMUTATION TO VALIDATE RE

```{r}
library(mgcv)

best_gam=mgcv::gam(LogY ~ s(Dipo,bs='cr')+s(ML,bs='cr')+s(EvDpt,bs='cr')+s(Station,bs="re")) ##where do I put nodes?
summary(best_gam)
ggplot.model(best_gam, res=F)
#best_gam$coefficients ##osservo gli effetti misti

#plot(best_gam$residuals,best_gam$fitted.values)
#qqnorm(best_gam$residuals)
#ks.test(x=best_gam$residuals,y='pnorm',alternative='two.sided') ###non gaussiano

best_gamb=mgcv::gam(LogY ~ s(Dipo,bs='cr')+s(ML,bs='cr')+s(EvDpt,bs='cr')) ##where do I put nodes?

print(anova(best_gam,best_gamb,test='F'))

print("AIC with RE")
AIC(best_gam)
print("AIC without RE")
AIC(best_gamb)

print("BIC with RE")
BIC(best_gam)
print("BIC without RE")
BIC(best_gamb)



#detach("package:mgcv", unload=TRUE)

```







